Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-19 article.Rmd
Stylistic changes
Stylistic changes to the section on Kantrowitz

The remainder of this article is divided into two sections. First, we describe our method in more detail, compare it to existing methods, and explain how to use it. Second, we apply the method to a case study of gatekeeping in the historical profession in order to demonstrate its utility.

@@

Why do scholars need a historical method for inferring gender? Let's start with a comparison. One existing method for predicting gender is available in the [Natural Language Toolkit](http://www.nltk.org/) for Python [Bird et al. 2009, Chapter 2].^[Chapter 6 of [Bird et al. 2009] also describes a method of predicting gender from the last letter of the name, but this example is given for the sake of explaining supervised classification and not as an end in itself, so we do not have reason to address this a historical method.] The NLTK is an influential software package for scholarship because it provides an extraordinary range of tools for analyzing natural language.^[[Bird et al. 2009] has been cited 812 times according to [Google Scholar](http://scholar.google.com/scholar?cites=3735380965977150091&as_sdt=5,47&sciodt=0,47&hl=en). Because in the humanities (unlike in other fields) the practice of citing software (even software which determines the outcome of an analysis) is regrettably rare, even nonexistent, such metrics are difficult to determine. Yet searching journals such as *Digital Humanities Quarterly* and the *Journal of Digital Humanities* reveals that natural language processing, for which the NLTK is the foremost implementation, is one of the key methodological categories of the digital humanities.] Included in the NLTK are two lists of male and female names created by the computer scientist Mark Kantrowitz with assistance from Bill Ross.^[The [original lists of names](http://www.cs.cmu.edu/afs/cs/project/ai-repository/ai/areas/nlp/corpora/names/0.html) are available at the website of the computer science department at Carnegie Mellon University, and the NLTK implementation is described in its [documentation](http://www.nltk.org/howto/corpus.html).] These lists include `r kantrowitz_unique` unique names. Using the Kantrowitz names corpus in the NLTK, one could look up a name like Jane or John and find out that it is male or female. The Kantrowitz corpus provides the list of names.^[The examples below is drawn from our Gender package which implements the Kantrowitz dataset for comparsion's sake, though its use is not recommended for actual research. The Gender package combines the Kantrowitz corpus's two lists into a single list, identifying names that are ambiguous because they appear on both lists as `"either"`.]

@@

The most significant problem with the Kantrowitz names corpus and thus the NLTK implementation is that it assumes that names are timeless. As pointed out above, this makes the corpus problematic for historical purposes. Furthermore the Kantrowitz corpus includes other oddities which make it less suitable. Some names like Abby are overwhelmingly female and some like Bill are overwhelmingly male, but the corpus includes them as both male and female. The Kantrowitz corpus contains only `r kantrowitz_unique` names, a mere `r percent(kantrowitz_unique / ssa_unique)` of the `r ssa_unique` names in the Social Security Administration dataset and `r percent(kantrowitz_unique / ipums_unique)` of the `r ipums_unique` names in the census records provided by Integrated Public Use Microdata (IPUMS) USA. There are therefore many names that it cannot identify. Assuming for the moment that our method provides more accurate results, we estimate that `r percent(kantrowitz_error_rate)` percent of the names in the Kantrowitz corpus are misclassified or classified as ambigious when a valid prediction could be made. **This isn't immediately obvious to me how this error rate was derived. Do we know why or how Kantrowitz is classifying things as ambiguous? Maybe he's just using a much higher, and theoretically more rigorous, cut-off (say 90%) than our 50% cut-off.**

>>>>

The remainder of this article is divided into two sections. First, we describe our method in more detail, compare it to existing methods, and explain how to use it. Second, we apply the method to a case study of gatekeeping in the historical profession in order to demonstrate its usefulness.

@@

Why do scholars need a historical method for inferring gender? Let's start with a comparison. One existing method for predicting gender is available in the [Natural Language Toolkit](http://www.nltk.org/) for the Python programming language [Bird et al. 2009, Chapter 2].^[Chapter 6 of [Bird et al. 2009] also describes a method of predicting gender from the last letter of the name, but this example is given for the sake of explaining supervised classification and not as an end in itself, so we do not have reason to address this as a historical method.] The NLTK is an influential software package for scholarship because it provides an extraordinary range of tools for analyzing natural language.^[[Bird et al. 2009] has been cited 812 times according to [Google Scholar](http://scholar.google.com/scholar?cites=3735380965977150091&as_sdt=5,47&sciodt=0,47&hl=en). Because in the humanities (unlike in other fields) the practice of citing software essential to performing scholarly analysis is regrettably rare, even nonexistent, such metrics are difficult to determine. Yet searching journals such as *Digital Humanities Quarterly* and the *Journal of Digital Humanities* reveals that natural language processing, for which the NLTK is the foremost implementation, is one of the key methodologies of the digital humanities.] Included in the NLTK are two lists of male and female names created by the computer scientist Mark Kantrowitz with assistance from Bill Ross.^[The [original lists of names](http://www.cs.cmu.edu/afs/cs/project/ai-repository/ai/areas/nlp/corpora/names/0.html) are available at the website of the computer science department at Carnegie Mellon University, and the NLTK implementation is described in its [documentation](http://www.nltk.org/howto/corpus.html).] These lists include `r kantrowitz_unique` unique names. Using the Kantrowitz names corpus in the NLTK, one could look up a name like Jane or John and find out whether it is male or female.

The Kantrowitz corpus provides the list of names.^[The examples below is drawn from our Gender package which implements the Kantrowitz dataset for sake of comparsion, though its use is not recommended for actual research. The Gender package combines the Kantrowitz corpus's two lists into a single list, identifying names that are ambiguous because they appear on both lists as `"either"`.]

@@

The most significant problem with the Kantrowitz names corpus and thus the NLTK implementation is that it assumes that names are timeless. As pointed out above, this makes the corpus problematic for historical purposes. Furthermore the Kantrowitz corpus includes other oddities which make it less useful for research. Some names such as Abby are overwhelmingly female and some such as Bill are overwhelmingly male, but the corpus includes them as both male and female. The Kantrowitz corpus contains only `r kantrowitz_unique` unique names, a mere `r percent(kantrowitz_unique / ssa_unique)` of the `r ssa_unique` unique names in a dataset provided by the Social Security Administration and `r percent(kantrowitz_unique / ipums_unique)` of the `r ipums_unique` unique names in the census records provided in the Integrated Public Use Microdata Series (IPUMS) USA dataset. There are therefore many names that it cannot identify. Assuming for the moment that our method provides more accurate results, we estimate that `r percent(kantrowitz_error_rate)` percent of the names in the Kantrowitz corpus are misclassified or classified as ambigious when a valid prediction could be made. **This isn't immediately obvious to me how this error rate was derived. Do we know why or how Kantrowitz is classifying things as ambiguous? Maybe he's just using a much higher, and theoretically more rigorous, cut-off (say 90%) than our 50% cut-off.**
